---
title: "CDT Module 1 - R Review"
author: "Valerie Bradley"
date: "6/10/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 99)
library(data.table)
```

###1. Sequences. 
Generate the following sequences using rep(), seq() and arithmetic:

(a) 1,3,5,7,...,21. 
```{r seq-a}
seq(1,21,2)
```
(b) 1,10,100,...,10^9.
```{r seq-b}
10^seq(0,9,1)
```
(c) 0,1,2,3,0,...,3,0,1,2,3 [with each entry appearing 6 times] 
```{r seq-c}
rep(seq(0,3,1),6)
```

(d) 0,0,0,1,1,1,2,...,4,4,4.
```{r seq-d}
sort(rep(seq(0,4,1),6))
```

(e) 50,47,44,...,14,11.
```{r seq-e}
seq(50,11,-3)
```

(f) 1,2,5,10,20,50,100,...,5×10^4.


###2. Arithmetic

(a) $cos \left( \frac{\pi n}{3} \right)$, for $n = 0,...,10$

```{r}
n = 0:10
cos(pi * n / 3)
```


(b) $1,9,98,997,...,999994$. 
```{r}
x = seq(0,6,1)
10^x - x
```
(c) $e^n-3n$,for $n=0,...,10$.
```{r}
n = 0:10
exp(1)^n - 3*n
```

(d) $3n$ mod $7$, for $n = 0,...,10$.
```{r}
n = 0:10
(3 * n) %% 7
```

(e) Let
$$
S_n = \sum_{i=1}^{n} \frac{(-1)^{i+1}}{2i-1} \\
\lim_{n \to \inf} Sn = \frac{\pi}{4}
$$
evaluate $4S_{10}$, $4S_{100}$ and $4S_{1000}$

```{r}
myseq = function(n){
  i = seq(1,n,1)
  4*sum((-1)^(i+1)/(2*i - 1))
}

data.frame(s_10 = myseq(10), s_100 = myseq(100), s_1000 = myseq(1000))
```

###3. Subsetting

```{r}
set.seed(123)
x = rnorm(100)
```

(a) the 25th, 50th and 75th elements; 
```{r}
x[c(25,50,75)]
```
(b) the first 25 elements;
```{r}
x[1:25]
```
(c) all elements except those from the 31st to the 40th.
```{r}
x[-(31:40)]
```


(d) all values larger than 1.5 (how many are there?);
```{r}
length(x[x>1.5])/100
1-pnorm(1.5)
```
(e) what about the entries that are either > 1.5 or < -1?
```{r}
x[x>1.5 | x < -1]
```

###4. Monte Carlo Integration
```{r}
z = rnorm(10000)
z_4 = z^4
mean(z_4)
hist(z_4)
```


```{r}
mean(z_4)

S_sq = 1 / (length(z_4) - 1) * sum((z_4 - mean(z_4))^2)
S_sq
# var(z_4) #check that it's the same

#get conf int
mean(z_4) + qnorm(c(0.025, 0.975)) * sqrt(S_sq)/sqrt(length(z_4))
```

###5. Records

```{r}
#x = rpois(n = 10000, lambda = 0.2)
x = rnorm(n = 10000, mean = 8, sd = 2)
temp = data.table(cbind(n = 1:length(x), x, cummax = cummax(x)))
temp[, record := (x == cummax)]
temp[record == TRUE,]

hist(temp[record == TRUE]$n)
```


