---
title: "CDT Module 1 - R Review"
author: "Valerie Bradley"
date: "9/19/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(data.table)
```

## 1. Vectors

a. Generate 100 standard normal random variables, and keep only the ones which are greater than 1. Don’t use a loop!
```{r cars}
all_norm = rnorm(100)
hist(all_norm)
hist(all_norm[all_norm > 1])
```

b. Write a function which takes two arguments n and min, and returns n independent random variables from a standard normal distribution truncated below by min. Let min default to 0.
```{r 1b}
rnorm_trunc = function(n, min = 0){
  X = rnorm(n)
  X = unlist(lapply(X, max, min))
  return(X)
}

hist(rnorm_trunc(2000, 1))
```


c. Generate 10k truncated normals with min set at -1 and plot as histogram
```{r}
samp = rnorm_trunc(10000, -1)
hist(samp, breaks  = 10)
```


d. what happens if min is large?
```{r}
hist(rnorm_trunc(10000,4))
```

Point here is that this method of rejection sampling is inefficient 


### 2. Data

```{r}
library(MASS)
data(hills)
summary(hills)
```

a. what sort of object is hills?
```{r}
class(hills)
```

b. how many columns?
```{r}
ncol(hills)
```

c. Change "Two Breweries" to "Three Breweries"
```{r}
hills[which(rownames(hills) == 'Two Breweries'),]
row.names(hills)[which(rownames(hills) == 'Two Breweries')] <-'Three Breweries'
hills[which(rownames(hills) == 'Two Breweries'),]
hills[which(rownames(hills) == 'Three Breweries'),]
```

d. Find the mean time for races iwth a climb greater than 1000ft
```{r}
with(hills[hills$climb > 1000,], mean(dist))
```

e. What sort of object is Orthodont?  How is it different from `hills`?
```{r}
library(nlme)
data(Orthodont)
head(Orthodont)
#class(Orthodont)
```
It's grouped data.


```{r}
head(methods(print))
#nlme:::print.groupedData(Orthodont)
```




### 3. Recursion

The $n$th Fibonacci number is defined by the recusion $F_n=F_{n-1} + F_{n-2}$ with $F_0=F_1=1$

a. Write a recursive function with argument `n` which returns the $n$th Fibonacci number

```{r}
fibonacci = function(n){
  if(n < 0) {
    #print(n)
    F_n = 0
  }else if(n <= 1){  #note this is different than the example given in Recall (they used <=2) - has to do with indexing from 0 v 1
    #print(n)
    F_n = 1
  }else{
    #print(n)
    F_n = Recall(n-1) + Recall(n-2)
  }
  F_n
}

fibonacci(2)   #evaluated 3 times

fibonacci(0)   #evaluated 1 time
fibonacci(1)   #evaluated 1 time
fibonacci(3)   #evaluated 5 times
```


```{r}
fib_2 = function(n){
  
  f_n_minus_1 = 1
  f_n_minus_2 = 1
  
  if(n < 0){
    return(0)
  }else if(n <= 1){
    return(1)
  }else{
    #initialize f_i at the first lag
    f_i = f_n_minus_1
    
     for(i in 2:n){
       
       #save old value
       f_i_old = f_i
       
       #get new value of f_i
       f_i = f_n_minus_1 + f_n_minus_2
       
       #update lag counters
       f_n_minus_2 = f_n_minus_1
       f_n_minus_1 = f_i_old
       
       print(c(i, f_i, f_n_minus_1, f_n_minus_2))
     }
    return(f_i)
  }
}
fib_2(1)
fib_2(2)
fib_2(3)
fib_2(4)
fib_2(5)
```


### 4. MCMC
a. 
X ~ Gamma(alpha, beta)
alpha, beta ~ Exp(1)

```{r}
#vector of data x
alpha_true = 1
beta_true = 2

x = rgamma(100, shape = alpha_true, rate = beta_true)


# evaluate the posterior distribution of alpha given a vector of data
posterior = function(x, alpha, beta){
  
  # get prior density
  prior_alpha = dexp(alpha, rate = 1, log = T)  # alpha ~ exp(1)
  prior_beta = dexp(beta, rate = 1, log = T)  # beta ~ exp(1)
  
  likelihood = dgamma(x, alpha, beta, log = T)
  
  posterior = prior_alpha + prior_beta + sum(likelihood)
  
  return(posterior)
}

posterior(x, alpha = 1.1, beta = 2.3)
posterior(x, alpha = 2, beta = 2)

```


b. single metropolis hasting step
$\alpha' = \alpha + \sigma Z_1$
$\beta' = \beta + \sigma Z_1$
where $Z_1 ,Z_2$ are iid Standard Normal $q'(\alpha' | \alpha) ~ N(\alpha, \sigma^2)$
```{r}
step_MH = function(x, alpha, beta, sigma){
  
  alpha_prime = rnorm(n = 1, alpha, sigma)
  beta_prime = rnorm(n = 1, beta, sigma)
  
  a = exp(posterior(x, alpha = alpha_prime, beta = beta_prime) - posterior(x, alpha = alpha, beta = beta))
  u = runif(1)
  if(u < a){
    data.frame(alpha = alpha_prime, beta = beta_prime)
  }else{
    data.frame(alpha = alpha, beta = beta)
  }
}

step_MH(x, alpha = 1.1, beta = 2.3, sigma = 0.01)

```

```{r}
run_MH = function(N, x, alpha, beta, sigma){
  chains = data.frame(t = 0, alpha = alpha, beta = beta)
  for(t in 1:N){
    step = step_MH(x, alpha = chains[t,]$alpha, beta = chains[t,]$beta, sigma)
    chains[t+1, ] = c(t, step$alpha, step$beta)
  }
  return(chains)
}
chains = run_MH(5000, x = x, alpha = 1.5, beta = 1, sigma = 0.1)


ggplot(chains, aes(x = t, y = alpha)) + geom_line(aes(color = 'alpha')) + geom_line(aes(x = t, y = beta, color = 'beta')) +
  ggtitle(paste0("Metropolis-Hastings Sampling for Gamma(alpha, beta)\n", "alpha = ", alpha_true, ", beta = ", beta_true)) + 
  geom_hline(yintercept = alpha_true) +
  geom_hline(yintercept = beta_true) +
  ylab("Parameter estimate")
```

```{r}
ggplot(chains) + stat_density(aes(x = alpha, fill = 'alpha', alpha = 0.1)) +
  stat_density(aes(x = beta, fill = 'beta', alpha = 0.1)) +
  geom_vline(xintercept = mean(chains$alpha), linetype = 'dotted') + 
  geom_vline(xintercept = alpha_true) +
  geom_vline(xintercept = mean(chains$beta), linetype = 'dotted') + 
  geom_vline(xintercept = beta_true) +
  ggtitle(paste0("M-H Sampling\nPosterior parameter distributions for Gamma(alpha, beta)\nT = ", nrow(chains) - 1))
```

Let's see how this converges over time
```{r}
Ts = seq(250, 5000, by = 250)

all_chains = lapply(Ts, FUN = run_MH, x = x, alpha = 1.5, beta = 1, sigma = 0.2)

all_chains = rbindlist(all_chains)
all_chains = as.data.table(all_chains)
all_chains[, N := rep(Ts, times = Ts + 1)]


all_chains[, .(mean(alpha), .N), N]
all_chains[, .(mean(beta), .N), N]

ggplot(all_chains[, .(alpha_hat = mean(alpha)), N]) + geom_point(aes(x = N, y = alpha_hat)) + geom_hline(yintercept = alpha_true)
ggplot(all_chains[, .(beta_hat = mean(beta)), N]) + geom_point(aes(x = N, y = beta_hat)) + geom_hline(yintercept = beta_true)

```



